Emotion Recognition in Written Conversations Using CASTLE

### 2021-01-27 msccls/MScTemplateCS.tex
report done!

%\geometry{showframe}
%better like this?
%\student{Flavius Gruian}{Flavius.Gruian@cs.lth.se}
\students{Nick Persson}{nick@persson.dev}{Filip Karabeleski}{fkarabeleski@gmail.com}

@@

We would like express our gratitude, first and foremost, to our supervisor Pierre. There is no way that this project would have been completed without your guidance, at least not in time.

@@

As the COVID-19 epidemic created some unusual circumstances, we would like to give a special thanks to \textit{Åke P Fågel \& Vilt} for providing us with a stable office to work out of. 
}

@@

%Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.
}

\keywords{Machine Learning, NLP, Emotion Recognition}

@@

\section{Background}
Natural language processing (NLP) is a field of study mainly in computer science that deals with analyzing human language in a way that computers can understand and process. According to \citet{ntlk2009}, natural language is used for everyday communication by humans, in contrast to artificial languages such as programming languages and mathematical notation. Since natural languages naturally evolve as time goes on and their adherence to structure diminish, it is difficult to define explicit rules to follow. For this reason, there is a need for machine learning models to aid the computer in identifying these diffuse rules.

In order to feed the model with effective information, the input is structured and inserted into a data set. This way, the computer has an easier way of processing and analyzing the data. When handling data sets constructed using exclusively textual input, the data sets are instead referred to as \textit{corpora}.

@@

\section{Purpose}
Customer service is centered around the relation between company and customer with the goal of pleasing the customer. This goal can be very difficult to achieve and there is no straightforward way to do so. By analyzing prior conversations, it is possible for companies to gain valuable insights in the most effective way to please the customer. Customer service is however a massive industry and it would take countless hours to analyze the conversations from even a single company manually.

The company sponsoring this project approached us with an idea to try and solve this problem using machine learning. A usual technique for these types of problems is \textit{sentiment analysis}, where utterances are labeled either as positive or negative. We instead decided to take another route, namely \textit{emotion recognition}. Emotion recognition focuses on labelling utterances from a non-binary set of emotions instead of sentiment analysis' binary method. This allows an analysis with greater specificity.

In order to further the academic relevance of this project we decided that we would attempt to contribute to the field of emotion recognition by achieving state-of-the-art results.

\section{State-of-the-art}
In this paper, the term state-of-the-art will be recurringly mentioned as it is a fundamental part of our aim and result. The term refers to the highest level of achievement accomplished. A common goal in machine learning projects is to attempt to raise the bar for the current state-of-the-art results across various metrics for the task that the project is trying to solve. Presenting state-of-the-art results is an excellent approach in proving that an implementation or algorithm is valid and valuable.

@@

\section{Telavox}
We carried out this project at Telavox AB. Telavox is a Swedish telecommunication company that specializes in business-to-business and business-to-customer unified communication as a service. As all of their customers deal in customer service, they would all benefit from the solution previously discussed. Telavox therefore has a great interest in being able to offer this as a service. As Telavox has over 250 businesses as customers, the solution would facilitate the quality of customer service world wide.

\section{Research questions}
\label{research_questions}
The primary aim of this thesis is to create a emotion recognition model that is able to produce state-of-the-art results. After this, a secondary aim is to investigate whether or not it is possible to create a dynamic model for instant response regarding customer service related conversations. The following questions are used as guidelines on how to realize these aims.

\begin{itemize}
    \item \textit{Is it possible to create a model that fulfills our aims?}
%?? THe first question is can we create such a model??
    \item \textit{Could a different architecture than the current state-of-the-art produce better results?}
%?? here your statement is too specific as you did not introduce the terms. Use vague words such as What are the different option to design an architecture and train a model?? 
%?? I have swapped the items ??
    \item \textit{How does our model compare to other models in the same field evaluating on the same metrics?}
    \item \textit{After implementing a fully functioning model, how can we adapt it to predict utterances in real-time?}
    %Snacka om detta i limitations
\end{itemize}

@@

In order to solve most emotion recognition problems, there are two main concepts that are fundamental to build an accurate model:
\begin{enumerate}
\item As shown by the recent progress in the field \citep{deep_survey}, the model should be created with deep learning algorithms. This should enable it to find patterns that it otherwise could not be able to. 
\item To effectively process the utterances, deep learning requires to collect an annotated corpus and design an appropriate representation for the words.
\end{enumerate}

@@

\section{Machine Learning}
Machine learning \citep{franoischollet2017learning} is a field of study in artificial intelligence that aims to train a system to solve a problem, rather than to explicitly program the system. By being presented with large amounts of data associated with the task at hand, the system algorithmically builds a statistical model that discovers structures in the data and allows the system to identify rules for the problem. 
In \textit{supervised learning} \citep{100pageBurkov}, which is one of the two main types of tasks in machine learning, the dataset is the collection of labeled examples $\{(\mathbf{x}_i, y_i)\}_{i=1}^N.$ Each element $\mathbf{x}_i$ is a feature vector that contains a value describing the example somehow and $y_i$ is a label, which is either a finite set of classes or a real number. The purpose of supervised learning is to use the dataset to produce a model that takes a feature vector $\mathbf{x}$ as input and outputs information that allows deducing the label $y$ for this feature vector.)

The other main type of task in machine learning is \textit{unsupervised learning}. The dataset for unsupervised learning is $\{ (\mathbf{x}_i)\}_{i=1}^N$ where $\mathbf{x}_i$ is a feature vector. The goal of unsupervised learning is to use the dataset to produce a model that takes a feature vector $\mathbf{x}$ as input and outputs either another vector or a value.

@@

Neural networks are machine learning systems loosely inspired by biological neural networks. A neural network is a collection of connected nodes or neurons that was first mentioned in \citet{mcculloch1943}. A neuron \citep{dawson1998ann} is an information-processing unit that receives multiple inputs and produces a single output that can be sent to multiple other neurons. To find the output of a neuron the weighted sum of all inputs is weighted by the weights all connecting to the neuron. This weighted sum is then passed through an activation function to provide the output.

@@

\begin{figure}[htp]
    \centering
    \includegraphics[width=12cm]{msccls/explanatory_images/neuron_new.png}
    \caption{An example of a single neuron. From \citet{wiki:neuron_image}.}
    \label{fig:neuron}
\end{figure}

@@

Neural networks can in turn be structured together and form multiple layers that progressively learn. These networks fall under the deep learning umbrella. Deep learning has several closely related definitions, where \citet{deng2014deep} mentions the following:

@@

Using the score produced by the loss function as feedback, the \textit{backpropagation} algorithm computes the \textit{gradient}, a multi-variable generalization of the derivative, which is then used to adjust the weights in such a way that the distance score produced by the loss function is minimized.

@@

One of the simplest neural network classes used in deep learning is \emph{recurrent neural networks} (RNN). RNN process either sequential data or time-series data \citep{lipton2015critical}. Since the inputs are sequential, all the inputs are dependent on each other in one way or another. RNNs are therefore suitable for processing sequences of text, audio, and video where every computation will be performed on each element in the sequence.

One of RNN's weaknesses is that they suffer from the \textit{vanishing gradient problem} \citep{hochreiter1998}. While RNN's theoretically can process arbitrarily long sequences, in practice, they are limited to the length of the sequence. If the sequence is too long, the network will struggle to try to look back in time longer than a few steps away. In a lot of cases, this will result in leaving out important information from the prediction. There are instead other alternatives that consider this weakness.

@@

 \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{msccls/explanatory_images/rnn.png}
    \caption{A simple RNN node}
    \label{fig:rnn_node}
\end{figure}

@@

As illustrated in Figure \ref{fig:lstm_node}, a LSTM node consists of an \textit{input gate}, an \textit{output gate}, a \textit{forget gate} and a \textit{cell state}. 
\begin{itemize}
    \item The input gate controls the extent of data to write onto the internal cell-state.
    \item The output gate controls the extent of data to pass as output. 
    \item The forget gate that decides how much of the previous data is irrelevant and will be ignored. 
    \item The cell state stores the previous state of a node and passes it on to its following node. 
\end{itemize}

@@

Basic GRU nodes  are primarily constructed of an \textit{update gate} and a \textit{reset gate}. The update gate regulates the amount of information that is passed along to the next state, while the reset gate, much like the forget gate of LSTM, decides the amount of past information that is going to be disregarded henceforth.

Although fairly new, it is already a well-established variant of recurrent neural networks. In comparison to \textit{LSTM} layers, GRU layers are cheaper to run since they use fewer parameters, but have less representational power. It has been shown by \citet{Gruber2020AreGC} that GRU layers have had better performance on smaller and less frequent datasets than LSTM layers.

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.58]{msccls/explanatory_images/lstm.png}
    \caption{A LSTM node} 
    %?? You need to mention the source of your figure if it is not yours??
    \label{fig:lstm_node}
\end{figure}

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.58]{msccls/explanatory_images/gru.png}
    \caption{A GRU node}
    \label{fig:gru_node}
\end{figure}

\section{Training Process}
In order to train a model, there must exist some data set, either labeled or unlabeled depending on whether supervised or unsupervised learning is the targeted method, that feeds the model with input. When using a labeled data set, the labels are regarded as the solution to the problem. When training a model, the goal is to find values for its weights such that the model correctly labels the given input.

The training of a model is done step by step in the form of \textit{epochs}. Epochs represent the number of times that the model iterates through a data set that is intended to train the model. Following each epoch, a \textit{loss function} measures how incorrect the model's previous prediction was and produces a value for the \textit{loss}. The model will have a perfect prediction when the loss reaches zero, therefore minimizing the loss is desirable.

The \textit{backpropagation} algorithm is a fundamental tool for minimizing the loss. The algorithm starts with the loss and works its way \textit{backwards} through the layers of the network and applies the chain rule to determine how much impact each of the weights has had on the loss. This impact is then used for the \textit{gradient descent} algorithm to alter the weights in the direction of flattening out the derivative. While travelling through the function, the ultimate goal is to find where the derivative is zero and this is done for each weight in the network. Gradient descent updates the weights to match the results from the backpropagation algorithm \citep{franoischollet2017learning}.

@@

\section{Evaluation}
Many different metrics can be used to evaluate a machine learning model, based on what kind of model and what its purpose is. For classification models, many of the metrics are based on a confusion matrix \citep{FAWCETT2006861}.

In a confusion matrix, one of the axes represents the actual class of the instance the model is trying to predict and the other represents what instance the model predicted. The most basic and common confusion matrix is a binary $2 \cdot 2$ matrix, as seen in Figure \ref{fig:confusion}. In binary cases, the instances usually can be represented as a yes or no question, such as \textit{is there a bird in this picture?}

@@

If we have a balanced dataset with a similar amount of positive and negative instances, then this can give valuable information on the performance of the model since getting predictions right is the goal of any model. If, however, our set is imbalanced and contains e.g. 90 percent negative instances and only 10 percent positive, the model can achieve a 90 percent accuracy by simply predicting negative on every instance so the accuracy suddenly becomes useless \citep{imbalancedlearning}.

\subsection{Precision}
Precision is a metric that quantifies how many of the instances predicted positive that were true. It's calculated by taking all of the instances where the model predicted positive and was correct and dividing it by all the instances where it predicted positive.

$$ Precision = \frac{TP}{TP + FP}.$$

Precision gives us information on how correct the model is when predicting an instance to be true. This is useful when dealing with instances, where it's important to not give a false positive. If we have a model that is trying to predict whether or not a suspect is guilty of a certain crime, a high precision would make it unlikely for an innocent person to be determined as a criminal.

@@

\subsection{Recall}
Recall is a metric that quantifies how many percent of positive instances that our model managed to predict which we calculate by dividing all instances the model predicted positive and was right by all positive instances.

$$ Recall =\frac{TP}{TP + FN}.$$

As a sort of counterpoint to precision which tells us how good the model is when guessing true, recall tells us how many positive instances the model missed. A high recall can be extremely important when dealing with subjects such as contagious diseases where a result of false-negative can be devastating.

Recall does not take into account how many positive predictions turned out to be false so in cases, where that does not matter and you want to predict positive at the slightest probability of it being true, optimizing for a high recall is essential.

@@

$$ F1score = 2 \cdot \frac{Precision \cdot Recall}{Precision + Recall}.$$

In cases where there is no extra cost associated with either false negatives or false positives, the goal is usually to minimize the total amount of errors. The harmonic mean is a type of average that is appropriate when working with ratios, which is what both precision and recall are.

This lets us improve the model concerning either of the variables and not have to think about if, for example, the reduction in Recall was worth the gain in Precision. In multi-class scenarios with imbalanced data sets there are multiple ways to calculate the F1-score, namely \textit{macro}, \textit{weighted} and \textit{micro}. Which one to use depends on your data set and what you prioritize in your model.

The macro F1-score is derived by calculating the F1-score for each class separately and then just summing them up, leading to an excessive focus on the classes with fewer instances. If we assume that we are using a data set where the \textit{neutral} class is represented in 10 000 instances and a less common emotion \textit{fear} in just 50. The model then managed to predict 9 995 of the neutral instances, but only 25 of the fear ones. When calculating the mean F1-score, the low fear score would muddy the entire score and make it seem bad, when in fact the score for the neutral case is excellent. If we want to optimize for the minority classes, however, this lets us do just that.

To counteract this behavior of favoring the minority class, one could instead calculate the weighted F1-score. After calculating the F1-score for each class, they are weighted by the number of true instances for each class, thus making them proportionate before calculating the final score. This gives a score more reflecting on how the model performs, but can also favor identifying the majority class heavily, which might not be the goal of the model.

When calculating micro F1 score, instead of calculating the F1 scores separately, the F1 score is calculated directly by using the total values for TP, FN, and FP in the Precision and Recall formulas which places no bias on either class. If we have an example data set with two minority classes \textit{anger} and \textit{joy}, we can see that for each instance where anger is true, classifying it as joy would be an FN from the perspective of the anger class. From the perspective of the joy class, however, it would instead be an FP. Since micro calculations take into account the total values for our identification labels and each FN correspond to the FP of another class, $FP = FN$, which in turn means that $Precision = Recall$ in these cases.

Precision and Recall being equal means that their mean also is equal and thus \textit{F1-}$score_{micro} = Precision_{Micro} = Recall_{Micro}$ which incidentally also equals \textit{Accuracy} as this is the proportion of instances we got correct. This means that just like Accuracy, this is a good metric on balanced data sets but can be deceiving in an imbalanced one.

\section{Natural Language Processing -- NLP}
Natural languages are the set of languages that humans have evolved naturally without any predetermined strategy such as English or Mandarin. Making computers understand and analyze natural languages has long been a challenge and one of the first practical attempts was Weaver and Booth's translation project in 1946 trying to crack enemy codes during World War II \citep{Liddy2001NLP}. Shortly thereafter \citet{Turing1950} created his famous \textit{Imitation game}, or more commonly referred to as the Turing test, which has developed into a test to evaluate whether a computer's natural language response in a real-time written conversation can be indistinguishable from that of a human.

One of the reasons that make natural language processing a challenge is the fact that natural languages are minted with unstructured types of data such as inflection or dialects, or using phrases to convey the opposite than the common meaning namely sarcasm. Natural languages are also oftentimes ambiguous and require a context-based interpretation that computers have a hard time understanding.

Emotion is a key part of human language and it, therefore, comes naturally that teaching a computer to understand emotion in human language is a big part of NLP. Initially, the focus lied in sentiment analysis, a field of study focused on determining the binary polarization of a piece of text as either positive or negative. 
%Godtycklig bit om sentiment analysis. 
Emotion recognition is an extension of sentiment analysis that instead of just determining the polarization of text, attributes emotion to it. This process is more challenging than sentiment analysis, as the differences between emotions are subtler than that of just positive and negative. 
Emotion recognition has also been extended to work on different modalities, such as visual and auditory input.

Instead of working directly with raw input, it first need to be converted into a numerical representation in order for the model to process it \citep{franoischollet2017learning}. In the case of textual input this is done by transforming it into a \textit{word vector}.

\subsection{One-Hot Encoding}
The most basic way of creating a word vector is the one-hot encoding method. This creates a very big and sparse $n \cdot n$ matrix where $n$ is the number of words in the embedding. Each word is then represented by an array filled with zeroes except for one index set to one. For example, the one-hot encoding of the phrase \textit{Why was six afraid of seven} would start with the following assignment of values to words:

As mentioned, this method is very basic and creates an unnecessarily large matrix that is largely filled up with nothing. If we wanted to represent a vocabulary of 20 000 words, each word would be represented by 19 999 zeroes and a single one.

The other big issue with it is that no meaning can be extracted from the matrix. If our vocabulary contained the words ``cat'', ``dog'' and ``hamster'', we could easily see that these are all domestic animals but in the space of the one-hot vector, ``cat'' could be as close to ``dog'' as it could be to ``airplane''.

\subsection{Word Embedding}
To resolve the two aforementioned problems with one-hot encodings, word embeddings were created, also known as dense word vectors \citep{neuralnetworkmethods}. 
Here, instead of creating a sparse, hardcoded vector with equal distance between words, they are learned from data and packed into a lower-dimensional vector where each dimension contains some relevant data about the word. According to \citet{franoischollet2017learning}, word embeddings that are 256-dimensional, 512-dimensional, or 1024-dimensional are common when dealing with very large vocabularies, in contrast to the 20 000-dimensional one that would be created if one were to use a one-hot encoding.

>>>>

%\geometry{showframe}
%better like this?
%\student{Flavius Gruian}{Flavius.Gruian@cs.lth.se}
\students{Filip Karabeleski}{fkarabeleski@gmail.com}{Nick Persson}{nick@persson.dev}

@@

We would like to express our gratitude, first and foremost, to our supervisor Pierre. There is no way that this project would have been completed without your guidance, at least not in time.

@@

As the COVID-19 pandemic created some unusual circumstances, we would like to give a special thanks to \textit{Åke P Fågel \& Vilt} for providing us with an office to use. 
}

@@

%Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.

Conversational emotion recognition is a notoriously challenging task for computers as conversations often contain subtle nuances that are difficult to identify. In this work, we investigated the recognition of conversational emotions in a set of dialogue corpora. We started from an existing baseline and, using empirical observation; we combined it with a previously untested set of architectures. We created a new framework that builds upon the previous state of the art and complements it with more effective recurrent devices. As a result, using this framework, CASTLE, we could improve the F1-score of previously published results on two of the three corpora we evaluated.}
% produces state-of the art results on several corpora commonly used as benchmarks for these types of tasks.
% possibility of improving the results in the field of

\keywords{NLP, emotion recognition, recurrent networks, deep learning}

@@

\section{Background}

Natural language processing (NLP) is a field of study mainly in computer science that deals with analyzing human language in a way that computers can understand and process. According to \citet{ntlk2009}, natural language is used for everyday communication by humans, in contrast to artificial languages such as programming languages and mathematical notation.

Concretely, NLP fields of application include machine translation, information extraction, and text categorization. For the last field, text categorization consists of automatically assigning a category to a text, for instance, to decide if a message is spam or not.

Since natural languages naturally evolve as time goes on and their adherence to structure diminishes, it is difficult to define explicit rules to follow. For this reason, there is a need for machine learning models to aid the computer in identifying these diffuse rules.

To build a machine-learning model, we need structured information as input in the form of a dataset. This way, the model will be able to learn parameters that are used to analyze and automatically categorize the data. Datasets using exclusively textual input are usually referred to as \textit{corpora}.

@@

\section{Purpose}
Customer service is centered around the relationship between company and customer with the goal of pleasing the customer. This goal can be challenging to achieve, and there is no straightforward way to do so. By analyzing prior conversations, companies can gain valuable insights in the most effective way to please the customer. However, customer service is a massive industry, and it would take countless hours to analyze the conversations from even a single company manually.

The company sponsoring this project approached us with an idea to try and solve this problem using machine learning. A usual technique for these problems is \textit{sentiment analysis}, where utterances are labeled either positive or negative. We instead decided to take another route, namely \textit{emotion recognition}. Emotion recognition focuses on labeling utterances from a non-binary set of emotions instead of sentiment analysis' binary method. This allows analysis with greater specificity.

To further the academic relevance of this project, we decided that we would attempt to contribute to the field of emotion recognition by achieving state-of-the-art results.

\section{State of the art}
%The progress of emotion recognition has been accomplished by 
%The previous work carried out in the field of emotion recognition has

In this report, the term \emph{state of the art} will be recurringly mentioned as it is a fundamental part of our aim and result. The term refers to the highest level of achievement accomplished. A common goal in machine learning projects is to attempt to raise the bar for the current
results across various metrics for the task that the project is trying to solve. Presenting state-of-the-art results is an excellent approach in proving that an implementation or algorithm is valid and valuable.

%Previous work

@@

\section{Telavox}
We carried out this project at Telavox AB. Telavox is a Swedish telecommunication company that specializes in business-to-business and business-to-customer unified communication as a service. As all of their customers deal in customer service, they would benefit from the previously discussed solution. Telavox, therefore, has a great interest in being able to offer this as a service. As Telavox has over 250 businesses as customers, the solution would facilitate the customer service quality worldwide.

\section{Research questions}
\label{research_questions}
This thesis' primary aim is to create an emotion recognition model that can produce state-of-the-art results. After this, a secondary aim is to investigate whether it is possible to create a dynamic model to provide interactive support in customer service-related conversations. The following questions are used as guidelines on how to realize these aims.

\begin{itemize}
    \item \textit{Is it possible to create a model that fulfills our aims?}
%?? THe first question is can we create such a model??
    \item \textit{Can we build an architecture that improves the state of the art?}
%?? here your statement is too specific as you did not introduce the terms. Use vague words such as What are the different option to design an architecture and train a model?? 
%?? I have swapped the items ??
    \item \textit{If yes, how does our model compare to other models in the same field evaluating on the same metrics?}
    \item \textit{After implementing a fully functioning model, how can we adapt it to predict utterances in real-time and in the context of an interactive service?}
    %Snacka om detta i limitations
\end{itemize}

@@

In order to solve most emotion recognition problems, two main concepts are fundamental to build an accurate model:
\begin{enumerate}
\item As shown by the recent progress in the field \citep{deep_survey}, the model should be created with deep learning algorithms. This should enable it to find patterns that it otherwise could not. 
\item To effectively process the utterances, deep learning requires collecting an annotated corpus and design an appropriate representation for the words.
\end{enumerate}

@@

\section{Machine Learning}
Machine learning \citep{franoischollet2017learning} is a field of study in artificial intelligence that aims to train a system to solve a problem rather than to program the system explicitly. By being presented with large amounts of data associated with the task at hand, the system algorithmically builds a statistical model that discovers structures in the data and allows the system to identify rules for the problem. 
In \textit{supervised learning} \citep{100pageBurkov}, which is one of the two main types of tasks in machine learning, the dataset is the collection of labeled examples $\{(\mathbf{x}_i, y_i)\}_{i=1}^N.$ Each element $\mathbf{x}_i$ is a feature vector that contains a value describing the example somehow, and $y_i$ is a label, which is either a finite set of classes or a real number. Supervised learning aims to use the dataset to produce a model that takes a feature vector $\mathbf{x}$ as input and outputs information that allows deducing the label $y$ for this feature vector.)

The other main type of task in machine learning is \textit{unsupervised learning}. The dataset for unsupervised learning is $\{ (\mathbf{x}_i)\}_{i=1}^N$, where $\mathbf{x}_i$ is a feature vector. The goal of unsupervised learning is to use the dataset to produce a model that takes a feature vector $\mathbf{x}$ as input and outputs either another vector or a value.

@@

Neural networks are machine learning systems loosely inspired by biological neural networks. A neural network is a collection of connected nodes or neurons first mentioned in \citet{mcculloch1943}. A neuron \citep{dawson1998ann} is an information-processing unit that receives multiple inputs and produces a single output sent to multiple other neurons. To find the output of a neuron, all inputs are weighted and then summed. This weighted sum is then passed through an activation function to provide the output.

@@

\begin{figure}[htp]
    \centering
    \includegraphics[width=12cm]{msccls/explanatory_images/neuron_new.png}
    \caption{Example of a single neuron. From \citet{wiki:neuron_image}.}
    \label{fig:neuron}
\end{figure}

@@

Neural networks can, in turn, be structured together and form multiple layers that progressively learn. These networks fall under the deep learning umbrella. Deep learning has several closely related definitions, where \citet{deng2014deep} mentions the following:

@@

Using the score produced by the loss function as feedback, the \textit{backpropagation} algorithm computes the \textit{gradient}, a multi-variable generalization of the derivative, which is then used to adjust the weights so that the distance score produced by the loss function is minimized.

@@

One of the simplest neural network classes used in deep learning is \emph{recurrent neural networks} (RNN). RNN process either sequential data or time-series data \citep{lipton2015critical}. Since the inputs are sequential, all the inputs are dependent on each other in one way or another. RNNs are suitable for processing sequences of text, audio, and video, where every computation will be performed on each element in the sequence.

One of RNN's weaknesses is that they suffer from the \textit{vanishing gradient problem} \citep{hochreiter1998}. While RNN's theoretically can process arbitrarily long sequences, they are, in practice, limited to the length of the sequence. If the sequence is too long, the network will struggle to try to look back in time longer than a few steps away. In many of cases, this will result in leaving out important information from the prediction. There are instead other alternatives that consider this weakness.

@@

 \begin{figure}[H]
    \centering
    \includegraphics[scale=0.35]{msccls/explanatory_images/rnn.png}
    \caption{Illustration of a simple RNN cell}
    \label{fig:rnn_node}
\end{figure}

@@

As illustrated in Figure \ref{fig:lstm_node}, an LSTM node consists of an \textit{input gate}, an \textit{output gate}, a \textit{forget gate}, and a \textit{cell state}. 
\begin{itemize}
    \item The input gate controls the extent of data to write onto the internal cell-state.
    \item The output gate controls the extent of data to pass as output. 
    \item The forget gate that decides how much of the previous data is irrelevant and will be ignored. 
    \item The cell state stores the previous state of a node and passes it on to its following node. 
\end{itemize}

@@

Basic GRU nodes  are primarily constructed of an \textit{update gate} and a \textit{reset gate}. The update gate regulates the amount of information passed along to the next state, while the reset gate, much like the forget gate of LSTM, decides the amount of past information that is going to be disregarded henceforth.

Although fairly new, it is already a well-established variant of recurrent neural networks. Compared to \textit{LSTM} layers, GRU layers are cheaper to run since they use fewer parameters but have less representational power. It has been shown by \citet{Gruber2020AreGC} that GRU layers have had better performance on smaller and less frequent datasets than LSTM layers.

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.58]{msccls/explanatory_images/lstm.png}
    \caption{Illustration of an LSTM cell} 
    %?? You need to mention the source of your figure if it is not yours??
    \label{fig:lstm_node}
\end{figure}

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.58]{msccls/explanatory_images/gru.png}
    \caption{Illustration of a GRU cell}
    \label{fig:gru_node}
\end{figure}

\section{Training Process}
To train a model, there must exist some dataset, either labeled or unlabeled depending on whether supervised or unsupervised learning is the targeted method that feeds the model with input. When using a labeled dataset, the labels are regarded as the solution to the problem. When training a model, the goal is to find values for its weights such that the model correctly labels the given input.

The training of a model is done step by step in the form of \textit{epochs}. Epochs represent the number of times 
the model iterates through a dataset intended to train the model. Following each epoch, a \textit{loss function} measures how incorrect the model's previous prediction was and produces a value for the \textit{loss}. The model will have a perfect prediction when the loss reaches zero, therefore minimizing the loss is desirable.

The \textit{backpropagation} algorithm is a fundamental tool for minimizing the loss. The algorithm starts with the loss and works its way \textit{backward} through the layers of the network, and applies the chain rule to determine how much impact each of the weights has had on the loss. This impact is then used for the \textit{gradient descent} algorithm to alter the weights to flatten out the derivative. While traveling through the function, the ultimate goal is to find where the derivative is zero, and this is done for each weight in the network. Gradient descent updates the weights to match the results from the backpropagation algorithm \citep{franoischollet2017learning}.

@@

\section{Natural Language Processing -- NLP}
One of the reasons that make natural language processing a challenge is that natural languages are minted with unstructured types of data such as inflection, dialects, or sarcasm. Natural languages are also oftentimes ambiguous and require a context-based interpretation that computers have a hard time understanding.

Emotion is a key part of human language, and therefore it comes naturally that teaching a computer to understand emotion in human language is a big part of NLP. Initially, the focus laid on sentiment analysis, a field of study focused on determining the binary polarization of a piece of text as either positive or negative. 
%Godtycklig bit om sentiment analysis. 
Emotion recognition is an extension of sentiment analysis that, instead of just determining text's polarization, attributes emotion to it. This process is more challenging than sentiment analysis, as the differences between emotions are subtler than that of just positive and negative. 
Emotion recognition has also been extended to work on different modalities, such as visual and auditory input.

Instead of working directly with raw input, it first needs to be converted into a numerical representation in order for the model to process it \citep{franoischollet2017learning}. In the case of textual input, this is done by transforming it into a \textit{word vector}.

\subsection{One-Hot Encoding}
The most basic way of creating a word vector is the one-hot encoding method. This creates a massive and sparse $n \cdot n$ matrix where $n$ is the number of words in the embedding. Each word is then represented by an array filled with zeroes except for one index set to one. For example, the one-hot encoding of the phrase \textit{Why was six afraid of seven} would start with the following assignment of values to words:

As mentioned, this method is elementary and creates an unnecessarily large matrix that is largely filled up with nothing. If we wanted to represent a vocabulary of 20 000 words, each word would be represented by 19 999 zeroes and a single one.

The other big issue with it is that no meaning can be extracted from the matrix. If our vocabulary contained the words ``cat'', ``dog'', and ``hamster'', we could easily see that these are all domestic animals, but in the space of the one-hot vector, ``cat'' could be as close to ``dog'' as it could be to ``airplane''.

\subsection{Word Embedding}
Word embeddings were created to resolve the two aforementioned problems with one-hot encodings, also known as dense word vectors \citep{neuralnetworkmethods}. 
Instead of creating a sparse, hardcoded vector with equal distance between words, they are learned from data and packed into a lower-dimensional vector where each dimension contains some relevant data about the word. According to \citet{franoischollet2017learning}, word embeddings that are 256-dimensional, 512-dimensional, or 1024-dimensional are common when dealing with extensive vocabularies, in contrast to the 20 000-dimensional one that would be created if one were to use one-hot encoding.

Figure \ref{fig:direction} shows the four words \textit{cat}, \textit{dog}, \textit{wolf}, and \textit{tiger} are represented on a geometric plane. We can clearly see some of the possible geometric transformations possible by using direction as a tool. For instance, the same vector that takes us from \textit{dog} to \textit{wolf} also takes us from \textit{cat} to \textit{tiger}. This could be interpreted this vector as ``converting a domesticated animal to its wildlife equivalent''.

Another vector will take us both from \textit{cat} to \textit{dog} and from \textit{tiger} to \textit{wolf}, which could be interpreted as ``converting a feline to a canine''. Two meaningful transformation vectors common in the real world are ``gender'' and ``plural'' vectors. Applying a ``female'' vector to \textit{king}, returns \textit{queen}, and the same vector returns \textit{actress} if instead used on \textit{actor}. ``Plural'', following the same logic, if used on \textit{actor} will return \textit{actors}.

%\subsection{RoBERTa}

\section{Evaluation}
Many different metrics can be used to evaluate a machine learning model, depending on what kind of model and what its purpose is. For classification models, many of the metrics are based on a confusion matrix \citep{FAWCETT2006861}.

In a confusion matrix, one of the axes represents the actual class of the instance the model is trying to predict, and the other represents what instance the model predicted. The most basic and common confusion matrix is a binary $2 \cdot 2$ matrix, as seen in Figure \ref{fig:confusion}. In binary cases, the instances usually can be represented as a yes or no question, such as \textit{is there a bird in this picture?}

@@

If we have a balanced dataset with a similar amount of positive and negative instances, this can give valuable information on the model's performance since getting predictions right is the goal of any model. If, however, our set is imbalanced and contains, e.g., 90 percent negative instances and only 10 percent positive, the model can achieve a 90 percent accuracy by simply predicting negative on every instance, so the accuracy suddenly becomes useless \citep{imbalancedlearning}.

\subsection{Precision}
Precision is a metric that quantifies how many of the instances predicted positive that were true. It is calculated by taking all of the instances where the model predicted positive and was correct and dividing it by all the instances where it predicted positive.

$$ Precision = \frac{TP}{TP + FP}$$

Precision gives us information on how correct the model is when predicting an instance to be true. This is useful when dealing with instances where it is important not to give a false positive. If we have a model trying to predict whether or not a suspect is guilty of a certain crime, a high precision will make it unlikely for an innocent person to be determined as a criminal.

@@

\subsection{Recall}
Recall is a metric that quantifies how many percent of positive instances that our model managed to predict, which we calculate by dividing all instances the model predicted positive and was right by all positive instances.

$$ Recall =\frac{TP}{TP + FN}$$

As a counterpoint to precision, which tells us how good the model is when guessing true, recall tells us how many positive instances the model missed. A high recall can be critical when dealing with e.g. contagious diseases where a result of false-negative can be devastating.

Recall does not consider how many positive predictions turned out to be false, so in cases where that does not matter, and you want to predict positive at the slightest probability of it being true, optimizing for a high recall is essential.

@@

$$ F1score = 2 \cdot \frac{Precision \cdot Recall}{Precision + Recall}$$

In cases where there is no extra cost associated with either false negatives or false positives, the goal is usually to minimize the total amount of errors. The harmonic mean is a type of average appropriate when working with ratios, which is what both precision and recall are.

This lets us improve the model concerning either of the variables and not have to think about if, for example, the reduction in Recall was worth the gain in Precision. There are multiple ways to calculate the F1-score in multi-class scenarios with imbalanced datasets, namely \textit{macro}, \textit{weighted} and \textit{micro}. Which one to use depends on your dataset and what you prioritize in your model.

The macro F1-score is derived by calculating the F1-score for each class separately and then summing them up, leading to an excessive focus on the classes with fewer instances. If we assume that we are using a dataset where the \textit{neutral} class is represented in 10 000 instances and a less common emotion \textit{fear} in just 50. The model then managed to predict 9 995 of the neutral instances, but only 25 of the fear ones. When calculating the mean F1-score, the low fear score would muddy the entire score and make it appear worse, when in fact, the score for the neutral case is excellent. If we want to optimize for the minority classes, however, this lets us do just that.

To counteract this behavior of favoring the minority class, one could instead calculate the weighted F1-score. After calculating the F1-score for each class, they are weighted by the number of true instances for each class, making them proportionate before calculating the final score. This gives a score more reflecting on how the model performs, but can also favor identifying the majority class heavily, which might not be the model's goal.

When calculating the micro F1 score, instead of calculating the F1 scores separately, the F1 score is calculated directly using the total values for TP, FN, and FP in the Precision and Recall formulas places no bias on either class. If we have an example dataset with two minority classes, \textit{anger}, and \textit{joy}, we can see that for each instance where anger is true, classifying it as joy would be an FN from the perspective of the anger class. From the perspective of the joy class, however, it would instead be an FP. Since micro calculations consider the total values for our identification labels and each FN correspond to the FP of another class, $FP = FN$, which means that $Precision = Recall$ in these cases.

Precision and Recall being equal means that their mean also is equal, and thus \textit{F1-}$score_{micro} = Precision_{Micro} = Recall_{Micro}$, which incidentally also equals \textit{Accuracy} as this is the proportion of instances we got correct. This means that just like Accuracy, this is a good metric on balanced datasets but can be deceiving in an imbalanced one.

\chapter{Implementation}
%Summary of what the chapter is about
This chapter is mainly focused on giving a clear description of what we have done for the project's duration. Firstly we will describe the model that we used as a baseline, \textit{COSMIC}. We will cover our model \textit{CASTLE}, its implementation, and what changes were made from COSMIC. Lastly, we will describe our setup and what corpora we used. 
%??In your text, you present COSMIC first. Change your intro so that it corresponds to your text??.
